---
title: Lecture 3 - Clustering
subtitle: OPTICS
bibliography: zotero.bib
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 3
---

```{r}
#| include: false
library(mclust)
library(ggplot2)
library(tibble)
library(dbscan)
```

# OPTICS: Ordering Points To Identify the Clustering Structure

## 1. Overview

- Definition:  
  OPTICS (Ordering Points To Identify the Clustering Structure) is a density-based clustering algorithm designed to handle clusters of varying densities.
  Unlike DBSCAN, which produces a flat partition, OPTICS creates an ordering of the data points along with their reachability distances.
  This ordering can then be visualized as a reachability plot.
  From here, clusters can be extracted based on the structure of the plot.

- Key idea:  
  Instead of committing to a single global density threshold, OPTICS records the structure of the data by computing for each point a reachability distance that reflects how it can be “reached” from a neighboring core point. Valleys in the reachability plot indicate dense clusters, and peaks mark boundaries between clusters.

- Main reference:  
  Ankerst, M., Brecheisen, M. M., Kriegel, H. P., Sander, J., & Xu, X. (1999). *OPTICS: Ordering Points To Identify the Clustering Structure*. SIGMOD Record, 28(2), 49–60.  @ankerst_etal99

## 2. Key Concepts

```{r}
#| include: false
path_img_optics_reachability <- projr::projr_path_get(
  "raw-img", "ankerst_etal99-reachability.png"
)
```

::: {#fig-optics-reachability}

![](`r path_img_optics_reachability`)

Core distance and reachability distance [@ankerst_etal99]
:::

### 2.1 Core distance

- For each point, the core distance is the minimum radius required to include at least `minPts` neighbors.
- If a point does not have enough neighbors, its core distance is undefined.

### 2.2 Reachability distance

- The reachability distance of a point $p$ with respect to a core point $o$ is defined as:
  
  $$
  r(p, o) = \max\bigl(\text{core-distance}(o),\, \text{distance}(o, p)\bigr)
  $$
  
- This measure shows how “reachable” $p$ is from $o$.

### 2.3 Reachability plot and cluster extraction

```{r}
#| include: false
path_img_optics_reachability_plot <- projr::projr_path_get(
  "raw-img", "ankerst_etal99-reachability-plot.png"
)
```

::: {#fig-optics-reachability-plot}
![](`r path_img_optics_reachability_plot`)

Reachability plot [@ankerst_etal99]
:::

- The OPTICS algorithm outputs an ordering of the data along with corresponding reachability distances.
- When plotted, the reachability plot shows valleys (dense regions) and peaks (gaps between clusters).
- Cluster extraction can be performed using methods such as the Xi method, which identifies steep changes in the plot to delimit clusters.

## 3. Applied to Data in R

```{r}
#| include: false
set.seed(123)
n_A <- 100
mean_A <- c(0, 0)
Sigma_A <- matrix(c(4, 0, 0, 1), nrow = 2)
data_A <- MASS::mvrnorm(n = n_A, mu = mean_A, Sigma = Sigma_A)
n_B <- 100
mean_B <- c(3, 3)
Sigma_B <- diag(2)
data_B <- MASS::mvrnorm(n = n_B, mu = mean_B, Sigma = Sigma_B)
data_sim <- rbind(data_A, data_B)
```

### 3.1 Running OPTICS

```{r}
# Run OPTICS with minPts = 10; eps is set high to include all relevant neighbors.
opt <- dbscan::optics(data_sim, minPts = 10, eps = 1.5)
```

### 3.2 Visualizing the Reachability Plot

```{r}
path_p_opt <- projr::projr_path_get(
  "cache", "notes-l3-cluster", "optics-reachability-plot.png"
)
png(path_p_opt, units = "cm", width = 10, height = 7, res = 300)
plot(opt)
dev.off()
```

This plot displays the reachability distance of each point (in the order they were processed).
Valleys indicate regions of high density that can be interpreted as clusters.

### 3.3 Extracting clusters using the DBSCAN method

```{r}
# Extract clusters using the Xi method with a specified xi threshold.
opt_clusters_dbscan <- dbscan::extractDBSCAN(opt, eps_cl = 0.6) 
```

```{r}
#| include: false
path_p_reachability_x_dbscan <- projr::projr_path_get(
  "cache", "notes-l3-cluster",
  "p-reachability-x-dbscan.png"
)
plot_tbl <- tibble(
  order = seq_along(opt_clusters_dbscan$reachdist),
  reachability = opt_clusters_dbscan$reachdist[opt_clusters_dbscan$order],
  cluster = as.character(opt_clusters_dbscan$cluster[opt_clusters_dbscan$order])
)

# plot a bar plot of reachability
# against the order
clust_col_vec <- RColorBrewer::brewer.pal(4, "Set1") |>
  setNames(
    nm = as.character(0:3)
  )
p <- ggplot(
  plot_tbl,
  aes(x = order, y = reachability, fill = cluster)
) +
  geom_bar(stat = "identity") +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    plot.background = element_rect(fill = "white")
  ) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank()
  ) +
  scale_fill_manual(
    values = clust_col_vec
  ) +
  labs(x = "Order", y = "Reachability")
ggsave(
  path_p_reachability_x_dbscan,
  plot = p,
  width = 8, height = 10, units = "cm", dpi = 300
)
```

::: {#fig-optics-reachability-x}

![](`r path_p_reachability_x_dbscan`)

Reachability plot with DBSCAN method cluster extraction
:::

```{r}
#| include: false
path_p_optics_dbscan <- projr::projr_path_get(
  "cache", "notes-l3-cluster",
  "p-optics-dbscan.png"
)
plot_tbl <- data_sim |>
  as_tibble() 
colnames(plot_tbl) <- c("x", "y")
plot_tbl$cluster <- as.character(opt_clusters_dbscan$cluster)

p <- ggplot(
  plot_tbl,
  aes(x = x, y = y, color = cluster)
) +
  geom_point() +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    plot.background = element_rect(fill = "white")
  ) +
  coord_equal() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank()
  ) +
  scale_colour_manual(
    values = clust_col_vec
  )
ggsave(
  path_p_optics_dbscan,
  plot = p,
  width = 8, height = 8, units = "cm", dpi = 300
)
```

::: {#fig-optics-xi}

![](`r path_p_optics_dbscan`)

OPTICS clustering using the DBSCAN method of noiseless data
:::

The extracted clusters (with labels) will correspond to valleys in the reachability plot as identified by the steepness threshold $\xi$.

### 3.3 Extracting clusters using the Xi method

```{r}
# Extract clusters using the Xi method with a specified xi threshold.
opt_clusters <- dbscan::extractXi(opt, xi = 0.04) # 0.04 caused the fuss
```

```{r}
#| include: false
path_p_reachability_x <- projr::projr_path_get(
  "cache", "notes-l3-cluster",
  "p-reachability-x.png"
)
plot_tbl <- tibble(
  order = seq_along(opt_clusters$reachdist),
  reachability = opt_clusters$reachdist[opt_clusters$order],
  cluster = as.character(opt_clusters$cluster[opt_clusters$order])
)
vline_tbl <- opt_clusters$clusters_xi |>
  as_tibble() |>
  tidyr::pivot_longer(
    cols = -cluster_id,
    names_to = "cluster",
    values_to = "order"
  ) |>
  dplyr::mutate(
    cluster_id = as.character(cluster_id)
  )
# plot a bar plot of reachability
# against the order
clust_col_vec <- RColorBrewer::brewer.pal(4, "Set1") |>
  setNames(
    nm = as.character(0:3)
  )
p <- ggplot(
  plot_tbl,
  aes(x = order, y = reachability, fill = cluster)
) +
  geom_bar(stat = "identity") +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    plot.background = element_rect(fill = "white")
  ) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank()
  ) +
  geom_vline(
    data = vline_tbl,
    aes(xintercept = order, color = cluster_id),
    linetype = "dashed",
    show.legend = FALSE
  ) +
  scale_fill_manual(
    values = clust_col_vec
  ) +
  scale_color_manual(
    values = clust_col_vec
  ) +
  labs(x = "Order", y = "Reachability")
ggsave(
  path_p_reachability_x,
  plot = p,
  width = 8, height = 10, units = "cm", dpi = 300
)
```

::: {#fig-optics-reachability-x}

![](`r path_p_reachability_x`)

Reachability plot with Xi method cluster extraction
:::

```{r}
#| include: false
path_p_optics_xi <- projr::projr_path_get(
  "cache", "notes-l3-cluster",
  "p-optics-xi.png"
)
plot_tbl <- data_sim |>
  as_tibble() 
colnames(plot_tbl) <- c("x", "y")
plot_tbl$cluster <- as.character(opt_clusters$cluster)

p <- ggplot(
  plot_tbl,
  aes(x = x, y = y, color = cluster)
) +
  geom_point() +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    plot.background = element_rect(fill = "white")
  ) +
  coord_equal() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank()
  ) +
  scale_colour_manual(
    values = clust_col_vec
  )
ggsave(
  path_p_optics_xi,
  plot = p,
  width = 8, height = 8, units = "cm", dpi = 300
)
```

::: {#fig-optics-xi}

![](`r path_p_optics_xi`)

OPTICS clustering using the XI method of noiseless data
:::

The extracted clusters (with labels) will correspond to valleys in the reachability plot as identified by the steepness threshold $\xi$.

### 3.4 Comparison with hierarchical clustering with single linkage

Single linkage hierarchical clustering gives less obvious clusters, likely due to the lack of penalisation of isolated points.

```{r}
#| include: false
dendro <- hclust(
  dist(data_sim),
  method = "single"
)
path_p_dendro <- projr::projr_path_get(
  "cache", "notes-l3-cluster",
  "p-optics-dendro.png"
)
png(path_p_dendro, units = "cm", width = 20, height = 20, res = 300)
plot(dendro)
dev.off()
```

::: {#fig-optics-dendro}

![](`r path_p_dendro`)

Dendrogram of single linkage hierarchical clustering
:::

However, the actual clustering result is probably the best we've had so far - if we label the first two clusters only and regard all others as noise (there were 20 clusters). 

```{r}
#| include: false
path_p_points_dendro_points <- projr::projr_path_get(
  "cache", "notes-l3-cluster",
  "p-optics-points-dendro.png"
)
plot_tbl <- data_sim |>
  as_tibble()
colnames(plot_tbl) <- c("x", "y")
clust_vec <- cutree(dendro, h = 0.59)
plot_tbl$cluster <- as.character(clust_vec)
p <- ggplot(
  plot_tbl,
  aes(x = x, y = y, color = cluster)
) +
  geom_point() +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    plot.background = element_rect(fill = "white")
  ) +
  coord_equal() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank()
  ) +
  scale_colour_manual(
    values = clust_col_vec[-length(clust_col_vec)]
  )
ggsave(
  path_p_points_dendro_points,
  plot = p,
  width = 8, height = 8, units = "cm", dpi = 300
)
```

::: {#fig-optics-points-dendro}

![](`r path_p_points_dendro_points`)

Classification using single linkage hierarchical clustering
:::

## 4. Parameters and Their Selection

### 4.1 minPts

- Definition:  
  The minimum number of points required to consider a point as a core point.  
- Selection:  
  Often set based on the dataset’s dimensionality (e.g., 10 for 2D data).
  Typically set higher than for DBSCAN to ensure robustness against noise.

### 4.2 eps

- Definition:  
  In OPTICS, `eps` is used as the maximum radius to search for neighbors.  
- Selection:  
  It should be large enough so that all points that might be reachable from any core point are considered.

### 4.3 Xi ($\xi$)

- Definition:  
  The parameter $\xi$ defines the relative steepness threshold in the reachability plot to detect significant changes (cluster boundaries).  
- Selection:  
  A smaller $\xi$ value (e.g., 0.05) yields more conservative cluster extraction, while larger values may merge clusters.

## 5. Advantages and Limitations

### 5.1 Advantages

- Handles varying densities:  
  Clusters of different densities can be identified (if using the Xi method).
- Noise detection:  
  Outliers appear as isolated points with high reachability distances.

### 5.2 Limitations

- Computational Cost:  
  OPTICS is generally more computationally intensive than DBSCAN because it computes a complete ordering of the data.
